[Re] Spike Synchronization and Rate Modulation Differentially Involved in Motor Cortical Function

### 2017-04-13 article/VRostami_JIto_MDenker_SGruen_2016.tex
fix typos and translation of Figure 5 to English

The UE analysis is performed in a sliding time window fashion, i.e.~the
data in each window are analyzed separately. This approach is chosen to
account for potential changes of the firing rates in time and to follow
the dynamics of the correlation. Here, a sliding time window is defined
based on trial time, meaning that a certain window position includes the
activity of all neurons in all trials in a certain time interval of the
trial. In the algorithm, at each position of the window, the data
contained in the window are extracted in order to compute the
significance of the specified patterns (line 3 in the pseudo code). For
doing that the UE analysis requires the empirical number of occurrences
of each pattern as well as its expected number given the firing rates.
While the empirical number is directly extracted from the data (line 5
in the pseudo code), the method to compute the expected number can be
chosen using the input parameter \emph{\texttt{method}} depending on the
assumptions regarding the data (line 6-19 in the pseudo code). For
cross-trial homogeneous data, selecting
\emph{``analytic\_TrialAverage''} as \emph{\texttt{method}} (line 7 in
the pseudo code) computes the expected number by estimating the rate of
each neuron from the average spike count across trials
\autocites{Gruen02a}{Gruen02b}. For data with cross-trial variability,
the \emph{``analytic\_TrialByTrial''} method (line 10 in the pseudo
code) should be used, which accounts for cross-trial changes in the
firing rates by computing the expected number of occurrences of the
spike pattern based on the product of single-trial estimates of firing
rates \autocite{Gruen03b}. Both methods perform a parametric test, where
the\\
\includegraphics{UE_algorithm.eps}\\
number of occurrences of the spike pattern is assumed to be a
Poisson-distributed. As an alternative non-parametric option, selecting
\emph{``surrogate\_TrialByTrial''} as \emph{\texttt{method}} (line 15 in
the pseudo code) will numerically compute the distribution of the
expected number by implementing the null-hypothesis based on surrogate
spike trains \autocite{Gruen09}. In the following, we will explain these
methods in greater detail.

@@

For the reproduction of the original results in \autocite{Riehle97}, we
have to focus on reproducing Figure 2A-F and Figure 4A, since for the
other figures the data still available were either incomplete in respect
to metadata (original Figure 3) or all the required data set were not
available anymore (original Figure 4B,C) . Figure 2 represents the main
result of the study and includes the UE analysis that underlies the
subsequent analyses. Figure 4A is an example of the application of the
UE method to data with more than 2 neurons. In terms of complexity of
the code the implementation of the UE analysis for three or more neurons
is considerably more demanding than for only two neurons. With this
example we show that our implementation is capable of performing the UE
analysis for the generic case of arbitrary numbers of neurons.

@@

Figures~\ref{fig:figure2alignedPS} and ~\ref{fig:figure2alignedRS} show
the results of performing the UE analysis for PS- and RS-aligned data,
respectively. Here, the analysis parameters are set to the identical
values as reported in the original publication (bin size: 5 ms, analysis
time window size: 100 ms, time step of the sliding window: 5 ms,
significance level \(\alpha= 0.05\)). The comparison of the two figures
to the original figure shows agreement in the raster displays (panel A)
and the time-resolved, trial-averaged firing rate estimates (B).
However, although the graphs of the number of coincidences per sliding
window (panel D) and the surprise measure (panel E) are similar in their
overall general behavior, they differ in the details. Thus, indeed the
choice of the alignment influences the analysis result. In order to test
if one of the two alignments is in agreement with the original
publication, we perform a detailed visual comparison of our two figures
and the original one on the basis of the spikes marked as coincident
(panels C) and as part of a UE (panel D). We notice that when aligning
to PS (Figure~\ref{fig:figure2alignedPS}), the marked spikes do not
agree in all details with the original figure. However, in
Figure~\ref{fig:figure2alignedRS}, with trials aligned to RS, we find no
disagreements with the original figure. After this visual comparison we
check if our results are exactly identical to the original ones.
Therefore, we extract the positions of the data points representing
spikes and UEs in the original figure by reading the plotting commands
in the PDF file of the original paper, and compare them to our
reproduced results. In Figure~\ref{fig:validation_fig2} A the reproduced
spike times in Figure~\ref{fig:figure2alignedRS} A are plotted against
the extracted original spike times. The plotted data points lie on the
diagonal line, indicating that the reproduced spike times correspond to
to the original ones. Figure~\ref{fig:validation_fig2} B shows the same
plot for the UEs, indicating the identity of the original and reproduced
UE timings as well. To further confirm the identity, we plot in
Figure~\ref{fig:validation_fig2} C the distributions of the differences
between the original and the reproduced timings for the spikes (black)
and the UEs (red). The differences are at most +/- 0.3 ms, which are
considerably narrower than the +/- 1 ms differences caused by the
misaligned data shown in Figure~\ref{fig:figure2alignedPS} A (see for
comparison the gray plot in Figure~\ref{fig:validation_fig2} C, which
shows the differences between the original spike times and the
misaligned spike times). The remaining minor differences of the spike
and UE times of the correctly aligned data and the original data are
only due to slight errors in the extraction of the spike times from the
original figure, which is inevitable because of a limited precision of
the plotting commands in the PDF file. Thus, we confirm that spike
timings in the correctly aligned data are identical to those in the
original data, and our implementation of the UE analysis applied to
these data reproduces exactly the same results as shown in the original
paper.

@@

\begin{figure}[htbp]
\centering
\includegraphics{CmpIDL_Matlab_3.eps}
\caption{\label{fig:workflow_report}Illustration of the data
preprocessing workflow, taken from the 1997 report of MD (in German) on
the comparison of the first UE implementation in IDL (left branch) and
the second implementation in Matlab \autocite{Diesmann16_personalcomm}
(right branch). Data entering both analysis branches have a resolution
of 0.1 ms (top box). In the IDL branch spike times \(t\) in the original
data are first transformed to a resolution of \(h=0.5\) ms (by the
\texttt{2gdf} program, middle left) using the method of binning
\(\lfloor t/h \rfloor\). Then the data are read into the ``IDL UE
Software'' (lower left box) and therein converted to \(h=1\) ms
resolution by the method of ``round half up''
\(\lfloor t/h+1./2 \rfloor\) prior to analysis. Alternatively, one can
load the 0.5 ms resolution data into the ``MATLAB UE Software'' (lower
right box). Here the bin width is a parameter of the analysis and thus
data can be converted to a 1 ms resolution but results are different
from the IDL branch. Results are only identical if a prior
transformation ``T'' (diagonal in center) performs a round half up and
no further binning is done in the MATLAB program. At a later point in
time the \texttt{alexa2gdf} converter function (written in MATLAB)
became available such that data in the original 0.1 ms resolution could
directly be converted to the 1 ms resolution by binning. The full report
(``Report\_by\_MarkusDiesmann.txt'') is included in the data folder of
the repository for this paper.}\label{fig:workflowux5freport}
\end{figure}

@@

We are able to reproduce the original results of \autocite{Riehle97} by
applying a new reimplementation of the Unitary Events analysis method in
Python to the original data. The method involves a number of numerical
computations and is very sensitive, as we show here by the comparison of
Figures ~\ref{fig:figure2alignedPS} and ~\ref{fig:figure2alignedRS},
which differed in the events that the trials were aligned to. This
difference in the alignment would not make any difference in the results
if the time difference between the two events (PS and RS) were identical
across the trials. But since the latter was not the case due to hardware
features of the recording setup (as we learned from the first author of
the original publication), the binning of the data started at a slightly
different time points in different trials. This likely led to a loss or
an addition of a spike in a bin and thus to a small difference of the
number of spike synchrony events (see also the discussion on the issues
of exclusive binning in \autocite{Gruen99}). In spite of this
sensitivity of the method, we succeeded in generating results that are
exactly identical to the original ones, as confirmed by the direct
comparison of the positions of spikes and UEs in the original and the
reproduced figures. This is a strong indication that our new
implementation of the analysis faithfully implemented the UE method.

@@

If the authors of the original paper would not have been accessible, we
would not have been able to reproduce the results. Nevertheless, here
our final validation of the reproduction is based the values extracted
from the vector graphic image (PDF) in the original paper. In an optimal
scenario, we would be able to exactly validate the results based on a
numerical comparison. To do so, all of the following pieces of
information would have had to be available at hand:

>>>>

The UE analysis is performed in a sliding time window fashion, i.e.~the
data in each window are analyzed separately. This approach is chosen to
account for potential changes of the firing rates in time and to follow
the dynamics of the correlation. Here, a sliding time window is defined
based on trial time, meaning that a certain window position includes the
activity of all neurons in all trials in a certain time interval of the
trial. In the algorithm, at each position of the window, the data
contained in the window are extracted in order to compute the
significance of the specified patterns (line 3 in the pseudo code). For
doing that the UE analysis requires the empirical number of occurrences
of each pattern as well as its expected number given the firing rates.
While the empirical number is directly extracted from the data (line 5
in the pseudo code), the method to compute the expected number can be
chosen using the input parameter \emph{\texttt{method}} depending on the
assumptions regarding the data (line 6-19 in the pseudo code). For
cross-trial homogeneous data, selecting
\emph{``analytic\_TrialAverage''} as \emph{\texttt{method}} (line 7 in
the pseudo code) computes the expected number by estimating the rate of
each neuron from the average spike count across trials
\autocites{Gruen02a}{Gruen02b}. For data with\\
\includegraphics{UE_algorithm.eps}\\
cross-trial variability, the \emph{``analytic\_TrialByTrial''} method
(line 10 in the pseudo code) should be used, which accounts for
cross-trial changes in the firing rates by computing the expected number
of occurrences of the spike pattern based on the product of single-trial
estimates of firing rates \autocite{Gruen03b}. Both methods perform a
parametric test, where the number of occurrences of the spike pattern is
assumed to be a Poisson-distributed. As an alternative non-parametric
option, selecting \emph{``surrogate\_TrialByTrial''} as
\emph{\texttt{method}} (line 15 in the pseudo code) will numerically
compute the distribution of the expected number by implementing the
null-hypothesis based on surrogate spike trains \autocite{Gruen09}. In
the following, we will explain these methods in greater detail.

@@

For the reproduction of the original results in \autocite{Riehle97}, we
have to focus on reproducing Figure 2A-F and Figure 4A, since for the
rest of the figures data were either incomplete in respect to metadata
(original Figure 3) or not available (original Figure 4B,C). Figure 2
represents the main result of the study and includes the UE analysis
that underlies the subsequent analyses. Figure 4A is an example of the
application of the UE method to data with more than 2 neurons. In terms
of complexity of the code the implementation of the UE analysis for
three or more neurons is considerably more demanding than for only two
neurons. With this example we show that our implementation is capable of
performing the UE analysis for the generic case of arbitrary number of
neurons.

@@

Figures~\ref{fig:figure2alignedPS} and ~\ref{fig:figure2alignedRS} show
the results of performing the UE analysis for PS- and RS-aligned data,
respectively. Here, the analysis parameters are set to the identical
values as reported in the original publication (bin size: 5 ms, analysis
time window size: 100 ms, time step of the sliding window: 5 ms,
significance level \(\alpha= 0.05\)). The comparison of the two figures
to the original figure shows agreement in the raster displays (panel A)
and the time-resolved, trial-averaged firing rate estimates (B).
However, although the graphs of the number of coincidences per sliding
window (panel D) and the surprise measure (panel E) are similar in their
overall general behavior, they differ in the details. Thus, indeed the
choice of the alignment influences the analysis result. In order to test
if one of the two alignments is in agreement with the original
publication, we perform a detailed visual comparison of our two figures
and the original one on the basis of the spikes marked as coincident
(panels C) and as part of a UE (panel D). We notice that when aligning
to PS (Figure~\ref{fig:figure2alignedPS}), the marked spikes do not
agree in all details with the original figure. However, in
Figure~\ref{fig:figure2alignedRS}, with trials aligned to RS, we find no
disagreements with the original figure. After this visual comparison we
check if our results are exactly identical to the original ones.
Therefore, we extract the positions of the data points representing
spikes and UEs in the original figure by reading the plotting commands
in the PDF file of the original paper, and compare them to our
reproduced results. In Figure~\ref{fig:validation_fig2} A the reproduced
spike times in Figure~\ref{fig:figure2alignedRS} A are plotted against
the extracted original spike times. The plotted data points lie on the
diagonal line, indicating that the reproduced spike times correspond to
the original ones. Figure~\ref{fig:validation_fig2} B shows the same
plot for the UEs, indicating the identity of the original and reproduced
UE timings as well. To further confirm the identity, we plot in
Figure~\ref{fig:validation_fig2} C the distributions of the differences
between the original and the reproduced timings for the spikes (black)
and the UEs (red). The differences are at most +/- 0.3 ms, which are
considerably narrower than the +/- 1 ms differences caused by the
misaligned data shown in Figure~\ref{fig:figure2alignedPS} A (see for
comparison the gray plot in Figure~\ref{fig:validation_fig2} C, which
shows the differences between the original spike times and the
misaligned spike times). The remaining minor differences of the spike
and UE times of the correctly aligned data and the original data are
only due to slight errors in the extraction of the spike times from the
original figure, which is inevitable because of a limited precision of
the plotting commands in the PDF file. Thus, we confirm that spike
timings in the correctly aligned data are identical to those in the
original data, and our implementation of the UE analysis applied to
these data reproduces exactly the same results as shown in the original
paper.

@@

\begin{figure}[htbp]
\centering
\includegraphics{CmpIDL_Matlab_3.eps}
\caption{\label{fig:workflow_report}Illustration of the data
preprocessing workflow, translated from the 1997 report of MD (in
German) on the comparison of the first UE implementation in IDL (left
branch) and the second implementation in Matlab
\autocite{Diesmann16_personalcomm} (right branch). Data entering both
analysis branches have a resolution of 0.1 ms (top box). In the IDL
branch spike times \(t\) in the original data are first transformed to a
resolution of \(h=0.5\) ms (by the \texttt{2gdf} program, middle left)
using the method of binning \(\lfloor t/h \rfloor\). Then the data are
read into the ``IDL UE Software'' (lower left box) and therein converted
to \(h=1\) ms resolution by the method of ``round half up''
\(\lfloor t/h+1./2 \rfloor\) prior to analysis. Alternatively, one can
load the 0.5 ms resolution data into the ``MATLAB UE Software'' (lower
right box). Here the bin width is a parameter of the analysis and thus
data can be converted to a 1 ms resolution but results are different
from the IDL branch. Results are only identical if a prior
transformation ``T'' (diagonal in center) performs a round half up and
no further binning is done in the MATLAB program. At a later point in
time the \texttt{alexa2gdf} converter function (written in MATLAB)
became available such that data in the original 0.1 ms resolution could
directly be converted to the 1 ms resolution by binning. The full report
(``Report\_by\_MarkusDiesmann.txt'') is included in the data folder of
the repository for this paper.}\label{fig:workflowux5freport}
\end{figure}

@@

We are able to reproduce the original results of \autocite{Riehle97} by
applying a new reimplementation of the Unitary Events analysis method in
Python to the original data. The method involves a number of numerical
computations and is very sensitive, as we show here by the comparison of
Figures ~\ref{fig:figure2alignedPS} and ~\ref{fig:figure2alignedRS},
which differed in the events that the trials were aligned to. This
difference in the alignment would not affect the results if the time
difference between the two events (PS and RS) were identical across the
trials. But since the latter was not the case due to hardware features
of the recording setup (as we learned from the first author of the
original publication), the binning of the data started at a slightly
different time points in different trials. This likely led to a loss or
an addition of a spike in a bin and thus to a small difference of the
number of spike synchrony events (see also the discussion on the issues
of exclusive binning in \autocite{Gruen99}). In spite of this
sensitivity of the method, we succeeded in generating results that are
exactly identical to the original ones, as confirmed by the direct
comparison of the positions of spikes and UEs in the original and the
reproduced figures. This is a strong indication that our new
implementation of the analysis faithfully implemented the UE method.

@@

If the authors of the original paper would not have been accessible, we
would not have been able to reproduce the results. Nevertheless, here
our final validation of the reproduction is based on the values
extracted from the vector graphic image (PDF) in the original paper. In
an optimal scenario, we would be able to exactly validate the results
based on a numerical comparison. To do so, all of the following pieces
of information would have had to be available at hand:
